Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic

### 2024-03-26 paper.Rmd
correcciones varias

---
title: "Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic"
css: "custom.css"
linestretch: '1.5'
link-citations: yes
# params:
# lang: en-GB
author: 
- name: Juan Carlos Castillo
  affiliation: Department of Sociology, Universidad de Chile
  email: juancastillov@uchile.cl
  number: 1, 3 
- name: Julio Iturra
  affiliation: Bremen International Graduate School of Social Sciences, Universität Bremen
  number: 2
- name: Kevin Carrasco
  affiliation: Center for Social Conflict and Cohesion Studies
  number: 3
abstract: | 
  Education is considered a key tool for social mobility and equality of opportunities. However, despite the widespread appreciation of education, disparities in educational outcomes persist over time, linked to variables such as families socioeconomic status, gender, ethnic origin, and migratory background, among others. While it is often assumed that there is a consensus on equal access to educational opportunities, empirical research on distributive justice teaches us that this assumption is far from reality. A significant body of evidence in this area indicates a substantial degree of acceptance and even justification of social inequalities, with education being no exception. In this regard, meritocracy emerges as one of the central concepts in the educational context related to justifying inequalities. As meritocracy emphasizes individual factors such as effort and talent in individual performance, the role of opportunities and social background is usually sidelined, leading to justifying inequalities of outcomes that perpetuate systemic disadvantages for marginalized groups. This research focuses on the changes in the justification of inequalities in access to education in Chile and how this justification is associated with the perception of meritocracy. The Chilean case offers an interesting context for this study due to its high economic inequality, deep neoliberal policies, and commodification of social services such as education. The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic). This change could be associated with a challenge of meritocratic ideals. Furthermore, we suggest that the association between a low meritocratic perception and less inequality justification would be stronger among traditionally vulnerable groups, such as individuals from lower social classes, less educated, and women. We analyze the ELSOC representative panel survey (2016 – 2022, 6 waves, N = 2,927), a study that includes various social, political, and economic topics during the last years. Estimating the statistical longitudinal models (multilevel random effects) shows mixed evidence about our hypotheses. Whereas we find support for the association between the perception of meritocracy and justification of inequality across time, the analyses that involved changes in justification of inequality are against our initial proposal. In addition, individuals in more advantaged situations and with greater resources tend to defend their interests and justify greater education inequality. Throughout the years, a larger perceived economic inequality has motivated a lower justification of inequality in education. 
  .
  \newline
  \newline

@@

# Acknowledgments

The authors recognize the support of ANID/FONDECYT grant 1160921; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046

@@

>>>>

---
title: "Changes in the justification of educational inequalities: The role of perceptions of inequality and meritocracy during the COVID pandemic"
css: "custom.css"
linestretch: '1.5'
link-citations: yes
# params:
# lang: en-GB
author: 
- name: Juan Carlos Castillo
  affiliation: Department of Sociology, Universidad de Chile
  email: juancastillov@uchile.cl
  number: 1, 3 
- name: Julio Iturra
  affiliation: Bremen International Graduate School of Social Sciences, Universität Bremen
  number: 2
- name: Kevin Carrasco
  affiliation: Center for Social Conflict and Cohesion Studies
  number: 3
abstract: | 
  Education is considered a key tool for social mobility and equality of opportunities. However, despite the widespread value of education, disparities in educational outcomes persist over time. To what extent are such educational disparities justified in society? And, what are the main factors driving this kind of justification? This research focuses on the changes in the justification of inequality in access to education in Chile and their links with the perception of meritocracy. The Chilean case offers an interesting context for this study given its high economic inequality, deep neoliberal policies, and commodification of social services such as education. The central argument of this article is that the justification of inequalities weakens during periods of vulnerability and crisis (such as the health and economic crisis resulting from the COVID-19 pandemic), which could be linked to a challenge of meritocratic ideals. For testing the research hypotheses we estimate a series of longitudinal multilevel models with data from a Chilean longitudinal panel survey (2016 – 2022, 6 waves, N = 2,927). Whereas we find support for the association between the perception of meritocracy and justification of inequality, the analyses that involved changes over time showed an increase in the justification of inequality in education. The discussion of these results delves into the social consequences of justifying inequality in a sensitive area as education, as well as the persistence of meritocratic ideals despite challenging events.

  .
  \newline
  \newline

@@

@@

_Acknowledgments_

The authors recognize the support of ANID/FONDECYT grant 1210847; the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046
